---
title: "Reproducible Research - Course Project 1"
author: "Shaun Bray"
date: "February 15, 2016"
output: html_document
---

# About

This is the first project in the Reproducible Research Course in the Data Science Specialization track. This project had us use Fitbit data and to answer question using that data.

# Data

The data for this assignment can be downloaded from the course website:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps takin in a 5-minute interval (missing values are coded as NA)
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

# Loading and preprocessing the data
Loading the data needed for analysis
```{r}
library(ggplot2)
library(plyr)
activity <- read.csv("activity.csv")
```
Processing the data
```{r}
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime <- as.POSIXct(activity$date,format="%Y-%m-%d")

# removes nas from data
cleandata <- activity[!is.na(activity$steps),]
```
# What is mean total number of steps taken per day?
Calculates the total number of steps taken per day.
```{r}
stepsbyday <- aggregate(activity$steps ~ activity$date, FUN=sum, )
colnames(stepsbyday) <- c("Date", "Steps")
```
Make a histogram of the total number of steps taken each day.
```{r}
hist(stepsbyday$Steps, breaks=5, xlab="Steps", main="Total Steps per Day")
```

Calculate and report the mean and median of the total number of steps taken per day.
```{r}
# Mean of steps
as.integer(mean(stepsbyday$Steps))
```
```{r}
# Median of steps
as.integer(median(stepsbyday$Steps))
```
The average number of steps taken eah day was 10,766.

The median number of steps taken each day was 10,765.

# What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).
```{r}
library(plyr)
library(ggplot2)
# Pulls data without nas
cleandata <- activity[!is.na(activity$steps),]

# Creates average number of steps per interval
interval_table <- ddply(cleandata, .(interval), summarize, Avg = mean(steps))

# Creates line plot of average number of steps per interval
p <- ggplot(interval_table, aes(x=interval, y=Avg), xlab = "Interval", ylab = "Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
# Maximum steps by interval
maxSteps <- max(interval_table$Avg)
# Which interval contains the maximum average number of steps
interval_table[interval_table$Avg==maxSteps,1]
```

The maximum number of steps for a 5-minute interval was 206 steps.

The 5-minute interval which had the maximum number of steps was the 835 interval.

# Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
# Number of NAs in original data set
nrow(activity[is.na(activity$steps),])
```

The total number of rows with steps = 'NA' is 2,304.

Devise a strategy for filling in all of the missing values in the data set. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

The strategy I will use for filling in the NAs is to substitute the missing steps with the average 5-minute interval based on the day of the week.
```{r}
# Create the average number of steps per weekday and interval
avg_table <- ddply(cleandata, .(interval, day), summarize, Avg = mean(steps))

# Create data set with all NAs for substitution
nadata <- activity[is.na(activity$steps),]
# Merge NA data with average weekday interval for substitution
newdata <- merge(nadata, avg_table, by=c("interval", "day"))
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
# Reorder the new substituded data in the same format as clean data set
newdata2 <- newdata[,c(6,4,1,2,5)]
colnames(newdata2) <- c("steps", "date", "interval", "day", "DateTime")

# Merge the NA averages and non NA data together
mergeData <- rbind(cleandata, newdata2)
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
# Create sum of steps per date to compare with step 1
stepsbyday2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(stepsbyday2) <- c("Date", "Steps")

# Mean of steps with NA data taken care of
as.integer(mean(stepsbyday2$Steps))

# Median of steps with NA data taken care of
as.integer(median(stepsbyday2$Steps))

# Creates the histogram of total steps per day
hist(stepsbyday2$Steps, breaks=5, xlab="Steps", main="Total Steps per Day with NAs Fixed", col="Black")
hist(stepsbyday$Steps, breaks=5, xlab="Steps", main="Total Steps per Day with NAs Fixed", col="Grey", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey"))
```

The new mean of the imputed data is `r mean(stepsbyday2$Steps)` steps compared to the old mean of `r mean(stepsbyday$Steps)` steps. That creates a difference of `r mean(stepsbyday2$Steps) - mean(stepsbyday$Steps)` steps on average per day.

The new median of the imputed data is `r median(stepsbyday2$Steps)` steps compared to the old median of `r median(stepsbyday$Steps)`. That creates a difference of `r median(stepsbyday2$Steps) - median(stepsbyday$Steps)` steps for the median.

However, the overall shape of the distribution has not changed.

# Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the data set with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
# Creates new category based on the days of the week
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
library(lattice)

# Summarize data by interval and type of day
interval_table2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))

# Plot data in a panel plot
xyplot(Avg~interval|DayCategory, data=interval_table2, type="l", layout = c(1,2), main="Average Steps per Interval Based on Type of Day", ylab="Average Number of Steps", xlab="Interval")
```

Yes, the activity trends are different based on whether the day occurs on a weekend or not. 
